# Geographic (GIS) data analysis 

In a couple of my projects, I have had the opportunity to visualize data using maps at the county level. Here is an example of the sort of maps I have made, using some simulated data. 

Suppose that a PI wants to study access to care through teledentistry. The specific research question is something like, "in a population of patients treated at my dental clinic from January 2021 - May 2023, what were the patterns in the relationships between 1) the distance traveled to the clinic, 2) the mode of the intake exam (in-person or virtual), and 3) treatment completion within 6 months (yes or no). Suppose further that I (the analyst) have access to the county in which each person lives, in addition to the clinical information relevant to our research in the electronic dental records. 

The first thing I would do is look online to find [publicly available data](https://geodata.iowa.gov/datasets/iowa::iowa-county-boundaries/explore?location=42.011193%2C-93.543500%2C8.00) with [Federal Information Processing Standard (FIPS) codes](https://en.wikipedia.org/wiki/FIPS_county_code) for Iowa at the county level. I will use the R package `sf` to handle GIS data, which often comes in `.shp` or `GeoJSON` formats. Data in these sorts of files have what we need to draw a map of Iowa with the counties demarcated on the map. 

```{r}
library(sf) # package for GIS data 
ia <- st_read("data/Iowa_County_Boundaries.geojson")
ia_geom <- st_geometry(ia)


# check - a test plot to make sure the map looks right. 
par(mar = c(0,0,1,0))
plot(ia[1], reset = FALSE) # reset = FALSE: we want to add to a plot with a legend
plot(ia[1,1], col = 'grey', add = TRUE)
```
This sample map is the correct shape -- here, the 99 counties are colored according to their index (1-99). We want to make a map where the colors correspond to the number of patients representing each county. 


Supposing one has access to electronic dental records, the data for such a study may look like this: 

```{r}
# simulate data
library(dplyr)
key <- read.csv("data/ia_counties.csv") |> 
  # narrow down to only IA (for sake of example)
  filter(State == "IA") # gives me county names 



# TD = teledentistry
set.seed(52242) 
td <- data.frame(
  id = 1:1000,
  year = sample(2021:2023, 1000, replace = TRUE),
  complete = sample(0:1, 1000, replace = TRUE),
  county = sample(key$NAME, 1000, replace = TRUE),
  mode = sample(c("virtual", "in-person"), 1000, replace = TRUE)
)

library(knitr)
head(td) |>
  kable(caption = "Simulated data from electronic dental records")
```

Now, I am ready to create a map that will communicate to my collaborators where our patients are driving from to receive their treatment. 

```{r}
# determine how many patients in each county
td_summarize <- td |>
    group_by(county) |>
    summarise(N = n()) %>%
  ungroup()

# add fips codes (from GeoJSON file)
td_summarize <- right_join(td_summarize, ia,
            by = c("county" = "CountyName")) 

# create sf object (for drawing a map)
map <- td_summarize %>%
  st_as_sf()

# draw the map 
library(ggplot2)
library(viridis)
ggplot() + 
  geom_sf(data = map,
          aes(fill = N)) + 
  scale_fill_viridis() + 
  labs(title = "Map of IA Residents by County", fill = "Number of \npatients") +
  theme_bw() 

```

